---
title: "Tonsil data clean up, mutaiton frequencies, BASELINe analysis and lineage reconstruction"
output: html_document
---

Bulk Ig RNA seq of B cells sorted into naïve (IgD+CD38-), GC (IgD-CD38+), plasma cells (IgD-CD38hi) and MBC (IgD-CD38-). 

# Load required packages
```{r load packages, message=FALSE, warning=FALSE}
library(alakazam)
library(shazam)
library(dplyr)
library(ggplot2)
library(tidyverse)
library(googledrive)
library(ggpubr)
library(Hmisc)
```

# Import the composite dataset
```{r import tonsil BCRs from all subjects, include=TRUE, warning=FALSE}
temp <- tempfile(fileext = ".rds")
dl <- drive_download(
as_id("11xEnLuFgpvRHxzYE00oTryXLOexn1y94"), path = temp, overwrite = TRUE)
bcp <- readRDS(file = dl$local_path)
```

# clean up subset and isotype information
```{r remove ambiguoius annotations, include=TRUE, warning=FALSE,message=FALSE}

bcp$germline <- sapply(strsplit(as.character(bcp$GERMLINE_V_CALL), split="*", fixed=TRUE), `[`, 1)

bcp <- bcp %>%
  subset(., SUBSET != "Total" & ISOTYPE %nin% c('IgE',
                                                'Primer6_Horns2016',
                                                'Primer1_1_70_Horns2016',
                                                'Primer4_1_Horns2016',
                                                'Primer2_Horns2016') &
           IN_FRAME == "TRUE")
```

# isolate VH4-34 sequences
```{r isoalte VH4-34, include=TRUE, message=FALSE, warning=FALSE}

V434 <- subset(bcp, germline=="IGHV4-34")
```

## clonality and lineage analysis
```{r split back into individual subjects and summarize expanded clones, include=TRUE, message=FALSE, warning=FALSE}

V434<-split(V434, f= V434$subject)

V434_clonesummary<-imap(V434, ~.x %>%
                            select(., c(SUBSET, CLONE_ID, ISOTYPE)) %>%
  group_by(SUBSET, CLONE_ID, ISOTYPE) %>%
  summarise(n = n()) %>%
    filter(., n>1 | duplicated(CLONE_ID)))

V434_clonesum_wide<-imap(V434_clonesummary, ~.x %>% 
                           pivot_wider(.,names_from = CLONE_ID, values_from = n))
```

# Process V434 dataframe to include only expanded clones and re-create clone summaries wider dataframe (keep clone_ids that are not unique)
```{r}
# all expanded clones
bcp434expall<-subset(V434, !isUnique(CLONE_ID))
bcp434expall<-split(bcp434expall, f= bcp434expall$subject)

bcp434expall_clonesummary<-imap(bcp434expall, ~.x %>%
                            select(., c(SUBSET, CLONE_ID, ISOTYPE)) %>%
  group_by(SUBSET, CLONE_ID, ISOTYPE) %>%
  summarise(n = n()))


bcp_clonesum_expall_wide<-imap(bcp434expall_clonesummary, ~.x %>% 
                          pivot_wider(.,names_from = CLONE_ID, values_from = n))

wb <- createWorkbook()
addWorksheet(wb, "VH434_allexpanded")

curr_row <- 1
for(i in seq_along(bcp_clonesum_expall_wide)) {
  writeData(wb, "VH434_allexpanded", names(bcp_clonesum_expall_wide)[i], startCol = 1, startRow = curr_row)
  writeData(wb, "VH434_allexpanded", bcp_clonesum_expall_wide[[i]], startCol = 1, startRow = curr_row+1)
  curr_row <- curr_row + nrow(bcp_clonesum_expall_wide[[i]]) + 2
}

saveWorkbook(wb, ".../VH434_all_expanded_clones_by_subset_isotype.xlsx")
```

# Import list of expanded clones
# create fucntion to import all sheets
```{r get target expanded clones from previous analysis, include=TRUE, message=TRUE, warning=FALSE}

library(readxl)     
multiplesheets <- function(fname) { # getting info about all excel sheets 
    sheets <- readxl::excel_sheets(fname) 
    tibble <- lapply(sheets, function(x) readxl::read_excel(fname, sheet = x)) 
    data_frame <- lapply(tibble, as.data.frame) 
    
    # assigning names to data frames 
    names(data_frame) <- sheets 
    
    # print data frame 
    print(data_frame) 
}

# specify the path
path<-".../06062024 CLONE IDs FOR LINEAGES WITH AVY MUTATIONS.xlsx"

cloneids<-multiplesheets(path)
clone3_4list<-cloneids$`All lineages 3-4`$...1
clone5morelist<-cloneids$`All lineages >5`$...1

bcp434_clones3_4<-subset(V434, CLONE_ID %in% clone3_4list)
bcp434_clones5more<-subset(V434, CLONE_ID %in% clone5morelist)
```

#Clones 3-4
#remove unequal length sequences
```{r remove sequences ith unequal lengths with germlin lignment, include=TRUE, message=FALSE, warning=FALSE}

#bcp434_clones3_4 <- data.frame(lapply(bcp434_clones3_4, as.character), stringsAsFactors=FALSE)
# then
x <- data.frame(bcp434_clones3_4$SEQUENCE_ID,
              nchar(as.character(bcp434_clones3_4$SEQUENCE_IMGT)))
y <- data.frame(bcp434_clones3_4$SEQUENCE_ID,
              nchar(as.character(bcp434_clones3_4$GERMLINE_IMGT)))
z <- cbind(x,y)
a <- subset(z, z$nchar.as.character.bcp434_clones3_4.SEQUENCE_IMGT.. !=z$nchar.as.character.bcp434_clones3_4.GERMLINE_IMGT..)

# and get the row numberss that have this discrepency, and remove them
bcp434_clones3_4_new <- subset(bcp434_clones3_4, 
                             bcp434_clones3_4$SEQUENCE_ID %nin%
                               a$bcp434_clones3_4.SEQUENCE_ID)

dim(bcp434_clones3_4)
dim(bcp434_clones3_4_new)
```

# clean up and rename columns so that they are recognized by Alakazam
```{r clean up column names, include=TRUE, message=FALSE, warning=FALSE}
bcp434_clones3_4 <- bcp434_clones3_4_new[!duplicated(bcp434_clones3_4_new$SEQUENCE_ID), ]
bcp434_clones3_4$locus <- "IGH"

bcp434_shortrees <- bcp434_clones3_4 %>%
                     dplyr::rename(., "sequence_id"="SEQUENCE_ID")%>%
                     dplyr::rename(., "clone_id"="CLONE_ID")%>%
                     dplyr::rename(., "sequence_alignment"="SEQUENCE_IMGT")%>%
                     dplyr::rename(., "germline_alignment"="GERMLINE_IMGT")%>%
                     dplyr::rename(., "v_call"="V_CALL")%>%
                     dplyr::rename(., "j_call"="J_CALL")%>%
                     dplyr::rename(., "junction_length"="JUNCTION_LENGTH")%>%
                     dplyr::rename(., "sequence"="SEQUENCE_INPUT")%>%
                     mutate(across(where(is.factor), as.character)) %>%
  mutate(dpucount= as.numeric(DUPCOUNT))

#sanity check
x <- data.frame(levels(factor(bcp434_shortrees$clone_id)))
```

# start creating individual clones because looping across the whole list doesnt work
```{r create individual trees four examples given here, include=TRUE, message=FALSE, warning=FALSE}

# BCP1_13463
BCP1_13463 <- subset(bcp434_shortrees, clone_id=="BCP1_13463")
BCP1_13463 <- makeChangeoClone(BCP1_13463, 
                             text_fields=c("sequence_id",  "ISOTYPE", "SUBSET"),
                             num_fields="DUPCOUNT",
                             pad_end=T)

BCP1_13463_clonetree <- buildPhylipLineage(BCP1_13463, phylip_exec=phylip_exec, rm_temp=TRUE)

#BCP1_13500
	 BCP1_13500 <- subset(bcp434_shortrees, clone_id=="BCP1_13500")
	 BCP1_13500 <- makeChangeoClone(BCP1_13500, 
                             text_fields=c("sequence_id",  "ISOTYPE", "SUBSET"),
                             num_fields="DUPCOUNT",
                             pad_end=T)
	 BCP1_13500_clonetree <- buildPhylipLineage(BCP1_13500, phylip_exec=phylip_exec, rm_temp=TRUE)
	 
#BCP1_27738
	 BCP1_27738 <- subset(bcp434_shortrees, clone_id=="BCP1_27738")
	 BCP1_27738 <- makeChangeoClone(BCP1_27738, 
                             text_fields=c("sequence_id",  "ISOTYPE", "SUBSET"),
                             num_fields="DUPCOUNT",
                             pad_end=T)
	 BCP1_27738_clonetree <- buildPhylipLineage(BCP1_27738, phylip_exec=phylip_exec, rm_temp=TRUE)
	 
#BCP1_35920
	 BCP1_35920 <- subset(bcp434_shortrees, clone_id=="BCP1_35920")
	 BCP1_35920 <- makeChangeoClone(BCP1_35920, 
                             text_fields=c("sequence_id",  "ISOTYPE", "SUBSET"),
                             num_fields="DUPCOUNT",
                             pad_end=T)
	 BCP1_35920_clonetree <- buildPhylipLineage(BCP1_35920, phylip_exec=phylip_exec, rm_temp=TRUE)
```

# concatenate shorter lineages into a list
```{r concatenate all trees into a list, include=TRUE, message=FALSE, warning=FALSE}
shortertrees<-grep("_clonetree",names(.GlobalEnv),value=TRUE)
shortertrees_list<-do.call("list",mget(shortertrees))
```

# Note: at this point, it is easier to create dataframes and plot trees individually/ manually
```{r plot tree and save tree information shown for one clone, include=TRUE, message=FALSE, warning=FALSE}

BCP1_13463_df <- data.frame(sequence_id=V(BCP1_13463_clonetree)$name, 
               ISOTYPE=V(BCP1_13463_clonetree)$ISOTYPE,subset=V(BCP1_13463_clonetree)$SUBSET,
               duplicate_count=V(BCP1_13463_clonetree)$DUPCOUNT,sequence=V(BCP1_13463_clonetree)$sequence)

wb <- openxlsx::createWorkbook()
addWorksheet(wb, "BCP1_13463")

writeData(wb, "BCP1_13463",BCP1_13463_df, startCol = 1, startRow = 1, colNames = F, rowNames = T)


# create tree
V(BCP1_13463_clonetree)$color[V(BCP1_13463_clonetree)$SUBSET == "Naive"] <- "orchid3"
V(BCP1_13463_clonetree)$color[V(BCP1_13463_clonetree)$SUBSET == "GC"] <- "red4"
V(BCP1_13463_clonetree)$color[V(BCP1_13463_clonetree)$SUBSET == "MBC"] <- "limegreen"
V(BCP1_13463_clonetree)$color[V(BCP1_13463_clonetree)$SUBSET == "PB"] <- "turquoise3"
V(BCP1_13463_clonetree)$color[V(BCP1_13463_clonetree)$name == "Germline"] <- "black"
V(BCP1_13463_clonetree)$color[grepl("Inferred", V(BCP1_13463_clonetree)$name)] <- "grey"
V(BCP1_13463_clonetree)$shape <- "none"
V(BCP1_13463_clonetree)$shape[V(BCP1_13463_clonetree)$ISOTYPE == "IgM"] <- "circle"
V(BCP1_13463_clonetree)$shape[V(BCP1_13463_clonetree)$ISOTYPE == "IgG"] <- "rectangle"
V(BCP1_13463_clonetree)$shape[V(BCP1_13463_clonetree)$ISOTYPE == "IgA"] <- "sphere"
V(BCP1_13463_clonetree)$shape[V(BCP1_13463_clonetree)$ISOTYPE == "IgA2"] <- "sphere"
par(mar=c(0, 0, 0, 0) + 0.05)


pdf(".../Tonsil data 2024/Trees/BCP1_13463_clonetree.pdf", width = 20, height = 10)

plot(BCP1_13463_clonetree, layout=layout_as_tree, vertex.frame.color="grey",
     vertex.label.color="black", edge.label.color="black",vertex.size=3,
     edge.arrow.mode=0)
# Add legend
legend("topleft", c("Germline", "Inferred", "Naive", "GC", "MBC", "PB"),
       fill=c("black", "grey", "orchid3", "red4", "limegreen", "turquoise3"), cex=0.75)
dev.off()


# After running all, save workbook
saveWorkbook(wb, ".../Tonsil data 2024/Trees/Shorter_lineage_tree_sequences_1-50.xlsx", overwrite = T)
```


# plot phylogenetic trees
```{r plot trees, include=TRUE, message=FALSE, warning=FALSE}

```

	  
## BASELINe analysis

# 1. Collapse isoypes to broad categories (IgM, IgG, IgA) and remove naive B cell IgG and IgA
```{r collapse Ig siotype subtypes in IgG and IgA, include=TRUE, message=FALSE, warning=FALSE}
VH434<-VH434%>%mutate(., iso =
                          case_when(ISOTYPE== "IgG1" ~ "IgG",
                                    ISOTYPE== "IgG2" ~ "IgG",
                                    ISOTYPE== "IgG3" ~ "IgG",
                                    ISOTYPE== "IgG4" ~ "IgG",
                                    ISOTYPE== "IgA1" ~ "IgA",
                                    ISOTYPE== "IgA2" ~ "IgA",
                                    ISOTYPE== "IgM" ~ "IgM",
                                    ISOTYPE== "IgD" ~ "IgD"))

# See how many class swithced naive B cells are present
naiv_igg<-subset(VH434, VH434$SUBSET=="Naive" & VH434$iso=="IgG")
nrow(naiv_igg)
naiv_iga<-subset(VH434, VH434$SUBSET=="Naive" & VH434$iso=="IgA")
nrow(naiv_iga)
# Remove IgG and IgA from naive B cells
VH434<-VH434[!(VH434$SUBSET=="Naive" & VH434$iso=="IgG"),] # No IGA but add a separate line of code, or add a | to separate the two conditions
```

# 2. Run CollapseClones
```{r collapse clones across the data, include=TRUE, message=FALSE, warning=FALSE}
db <- collapseClones(VH434, cloneColumn="CLONE_ID",
sequenceColumn="SEQUENCE_IMGT",
germlineColumn="GERMLINE_IMGT_D_MASK",
method="thresholdedFreq", minimumFrequency=0.6,
includeAmbiguous=FALSE, breakTiesStochastic=FALSE)
```

# 3. Sepearte db into IgM, IgG and IgA
```{r separate the data into isotypes, include=TRUE, warning=FALSE, message=FALSE}
db_igm<-subset(db, db$iso=="IgM")
db_igg<-subset(db, db$iso=="IgG")
db_iga<-subset(db, db$iso=="IgA")
```

# 4.a. IgM analysis
```{r BASELINe IgM analysis, include=TRUE, message=FALSE, warning=FALSE}
baseline_region_igm <- calcBaseline(db_igm,
                               sequenceColumn="clonal_sequence",
                               germlineColumn="clonal_germline",
                               testStatistic="focused",
                               regionDefinition=IMGT_V_BY_REGIONS,
                               targetingModel=HH_S5F,
                               nproc=1)
grp_igm<-groupBaseline(baseline_region_igm, groupBy="SUBSET", nproc = 1)
```

# 4.b. IgG analysis
```{r BASELINe IgG analysis, include=TRUE, message=FALSE, warning=FALSE}
baseline_region_igg <- calcBaseline(db_igg,
                                    sequenceColumn="clonal_sequence",
                                    germlineColumn="clonal_germline",
                                    testStatistic="focused",
                                    regionDefinition=IMGT_V_BY_REGIONS,
                                    targetingModel=HH_S5F,
                                    nproc=1)
grp_igg<-groupBaseline(baseline_region_igg, groupBy="SUBSET", nproc = 1)
```

# 4.c. IgA analysis
```{r BASELINe IgA VH434, include=TRUE, message=FALSE, warning=FALSE}
baseline_region_iga <- calcBaseline(db_iga,
                                    sequenceColumn="clonal_sequence",
                                    germlineColumn="clonal_germline",
                                    testStatistic="focused",
                                    regionDefinition=IMGT_V_BY_REGIONS,
                                    targetingModel=HH_S5F,
                                    nproc=1)
grp_iga<-groupBaseline(baseline_region_iga, groupBy="SUBSET", nproc = 1)
```

# 5. Plot probability densities for each isotype
```{r Plot probability densities for each isotype, include=TRUE, warning=FALSE, message=FALSE}
subset_colors <- c("Naive"="darkorchid", "MBC"="firebrick",
"GC"="seagreen", "PB"="steelblue")
# And plot plots
# IgM
igm_cdr1<-plotBaselineDensity(grp_igm,  "SUBSET", 
                              colorValues=subset_colors,
                              subsetRegions="cdr1",
                              sigmaLimits=c(-2, 2))
igm_cdr2<-plotBaselineDensity(grp_igm,  "SUBSET", 
                              colorValues=subset_colors,
                              subsetRegions="cdr2",
                              sigmaLimits=c(-2, 2))
igm_fwr1<-plotBaselineDensity(grp_igm,  "SUBSET", 
                              colorValues=subset_colors,
                              subsetRegions="fwr1",
                              sigmaLimits=c(-2, 2))
igm_fwr2<-plotBaselineDensity(grp_igm,  "SUBSET", 
                              colorValues=subset_colors,
                              subsetRegions="fwr2",
                              sigmaLimits=c(-2, 2))
igm_fwr3<-plotBaselineDensity(grp_igm,  "SUBSET", 
                              colorValues=subset_colors,
                              subsetRegions="fwr3",
                              sigmaLimits=c(-2, 2))

#IgG
igg_cdr1<-plotBaselineDensity(grp_igg,  "SUBSET", 
                              colorValues=subset_colors,
                              subsetRegions="cdr1",
                              sigmaLimits=c(-2, 2))
igg_cdr2<-plotBaselineDensity(grp_igg,  "SUBSET", 
                              colorValues=subset_colors,
                              subsetRegions="cdr2",
                              sigmaLimits=c(-2, 2))
igg_fwr1<-plotBaselineDensity(grp_igg,  "SUBSET", 
                              colorValues=subset_colors,
                              subsetRegions="fwr1",
                              sigmaLimits=c(-2, 2))
igg_fwr2<-plotBaselineDensity(grp_igg,  "SUBSET", 
                              colorValues=subset_colors,
                              subsetRegions="fwr2",
                              sigmaLimits=c(-2, 2))
igg_fwr3<-plotBaselineDensity(grp_igg,  "SUBSET", 
                              colorValues=subset_colors,
                              subsetRegions="fwr3",
                              sigmaLimits=c(-2, 2))

#Iga
iga_cdr1<-plotBaselineDensity(grp_iga,  "SUBSET", 
                              colorValues=subset_colors,
                              subsetRegions="cdr1",
                              sigmaLimits=c(-2, 2))
iga_cdr2<-plotBaselineDensity(grp_iga,  "SUBSET", 
                              colorValues=subset_colors,
                              subsetRegions="cdr2",
                              sigmaLimits=c(-2, 2))
iga_fwr1<-plotBaselineDensity(grp_iga,  "SUBSET", 
                              colorValues=subset_colors,
                              subsetRegions="fwr1",
                              sigmaLimits=c(-2, 2))
iga_fwr2<-plotBaselineDensity(grp_iga,  "SUBSET", 
                              colorValues=subset_colors,
                              subsetRegions="fwr2",
                              sigmaLimits=c(-2, 2))
iga_fwr3<-plotBaselineDensity(grp_iga,  "SUBSET", 
                              colorValues=subset_colors,
                              subsetRegions="fwr3",
                              sigmaLimits=c(-2, 2))
```
